Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-10-30 paper/analysis.tex
add citations of recently published preprint

\color{blue}
In this section, we consider a general parameter-dependent matrix
\begin{equation*}
    \mtx{B}(t) = \begin{bmatrix}
        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n},
\end{equation*}
where each entry $b_{ij}(t)$ is a continuous function on a bounded interval $[a,b]$. \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
$\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$. This becomes more difficult when $\mtx{B}(t)$ is given implicitly (e.g., as a matrix function) and can only be accessed through matrix-vector products. The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors

as we will see in Section ???.
\color{black}

@@

\subsubsection{Nyström approximation}

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then for any $\gamma \geq 1$ the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
        a universal constant $c > 0$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

>>>>

\color{blue}
In this section, we consider a general parameter-dependent matrix
\begin{equation*}
    \mtx{B}(t) = \begin{bmatrix}
        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n},
\end{equation*}
where each entry $b_{ij}(t)$ is a continuous function on a bounded interval $[a,b]$. \emph{If} these entries are given explicitly, we can simply compute the trace by its definition 
$\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$. This becomes more difficult when $\mtx{B}(t)$ is given implicitly (e.g., as a matrix function) and can only be accessed through matrix-vector products. The common theme of all methods analyzed in this section is to extract trace estimates from multiplying $\mtx{B}(t)$ with fixed random vectors.

%as we will see in Section ???.
\color{black}

@@

\subsubsection{Nyström estimator}

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then for any $\gamma \geq 1$ the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
        $c = 1 + 9 \sqrt{2} e^4$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

### 2024-10-30 paper/intro.tex
add citations of recently published preprint

Spectral distributions can reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations eigenvalues represent the energy levels which electrons occupy\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape \cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.

Being composed of distributions, the function $\phi$ itself is hard to approximate. However, as many application only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A})),
    \label{equ:smoothed-spectral-density}
\end{equation}
for some fixed smoothing kernel $g_{\sigma}$, typically a Gaussian~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian~\cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. We thus arrive at the problem of computing the trace of the parameter-dependent matrix function $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Evaluating this matrix function \emph{exactly} requires the diagonalization of $\mtx{A}$, which does not seem to yield any gains compared to the original problem of computing its spectral distribution. The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $\mtx{A}$.

>>>>

Spectral distributions can reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations eigenvalues represent the energy levels which electrons occupy\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape \cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2024-improved-spectral}.

Being composed of distributions, the function $\phi$ itself is hard to approximate. However, as many application only require a rough estimate of $\phi$, it usually suffices to instead approximate a smoothed spectral density of the form
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A})),
    \label{equ:smoothed-spectral-density}
\end{equation}
for some fixed smoothing kernel $g_{\sigma}$, typically a Gaussian~\cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. We thus arrive at the problem of computing the trace of the parameter-dependent matrix function $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Evaluating this matrix function \emph{exactly} requires the diagonalization of $\mtx{A}$, which does not seem to yield any gains compared to the original problem of computing its spectral distribution. The crucial point is that we can now \emph{estimate} the trace by matrix-vector products with $\mtx{B}(t)$, which -- thanks to the smoothness of $g_\sigma$ -- can be well approximated by a few matrix-vector products with $\mtx{A}$.

### 2024-10-30 paper/spectraldensity.tex
add citations of recently published preprint

Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}. To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
\begin{equation}
    \sup_{f \in \mathcal{S}} \int_{a}^{b} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
    \label{equ:error-metric}
\end{equation}
where $\mathcal{S}$ represents an appropriately chosen space of test functions. Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time} uses $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$. In this last metric, provided we assume the eigenvalues to be somewhat uniformly distributed and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$. Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.

@@

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \DCT(\vct{\mu})$
    \State Compute $\vct{\nu} = \DCT^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.

>>>>

Choosing the smoothing parameter $\sigma$ results in the following trade off: Typically, larger $\sigma$ make it easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas smaller $\sigma$ allow us to stay closer to the original spectral density \refequ{equ:spectral-density}. To see the latter, we measure the error between the spectral density $\phi$ and its smoothened version $\phi_{\sigma}$ with
\begin{equation}
    \sup_{f \in \mathcal{S}} \int_{a}^{b} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
    \label{equ:error-metric}
\end{equation}
where $\mathcal{S}$ represents an appropriately chosen space of test functions. Among others, \cite{lin-2016-approximating-spectral} proposes $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ where $\Theta$ is the Heaviside step function, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$. In this last metric, provided we assume the eigenvalues to be somewhat uniformly distributed and $\sigma$ to be rather small, a simple calculation allows us to roughly estimate the smoothing error \refequ{equ:error-metric} to be $\sigma$. Consequently, if we can only allow an approximation to deviate by at most a factor of $\varepsilon > 0$ from the original spectral density, we should choose a smoothing parameter $\sigma \approx \varepsilon$.

@@

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.
